---
title: "Storm Data Project"
course: "Reproducible Research - Assignment 2"
data: National Oceanic and Atmospheric Administration (NOAA)
author: "KB"
date: "Thursday, November 19, 2015"
output: html_document
---

#Reproducible Research - Assessment 2

#Title : Analysis of NOAA Storm Database

##Synopsis

This analysis consist of exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database,
which tracks characteristics of major storms and weather events in the United States, including when and where they
occur, as well as estimates of any fatalities, injuries, and property damage.
Storms and other severe weather events can cause both public health and economic problems for communities and 
municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such
outcomes to the extent possible is a key concern.
The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.
This analysis tries to answer the following questions.

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

##Data Processing

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.
- National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
- National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

### Loading and Reading Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from this link:
- [Storm Data (47Mb)](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

```{r}
library(dplyr)
library(ggplot2)
if(!file.exists("./repdata-data-StormData.csv.bz2")){
    temp<-tempfile()
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                temp)
    remove(temp)
    }
Storm_data <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))
```

### Exploraing the Data

Data consists of 902,207 rows and 37 columns collected during 62 years

```{r, echo=TRUE}
dim(Storm_data)
#summary(Storm_data)
str(Storm_data)
head(Storm_data)
#Subsetting the data to show only those columns we wneed to work on
onlyColumns <- c("BGN_DATE","COUNTY","COUNTYNAME","STATE","EVTYPE","LENGTH","WIDTH","FATALITIES","INJURIES","PROPDMG", "PROPDMGEXP","CROPDMG","CROPDMGEXP")
StormData <- Storm_data[, onlyColumns]
StormData$YEAR <- as.numeric(format(as.Date(Storm_data$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
#unique(StormData$YEAR)
TotStormsPerYear <- split(as.numeric(Storm_data$fatalities), Storm_data$year)
#missingvalues<-sum(is.na(Storm_data))
```

###Estimating Property damage by dollar amount

Property damage estimates should be entered as actual dollar amounts. To prepare the data, we need to assign a dollar amount value according with the magnitude of the damage.

```{r}
#exploring Property damage data
unique(StormData$PROPDMGEXP)
# Sorting the property exponent data
StormData$PROPEXP[StormData$PROPDMGEXP == "K"] <- 1000
StormData$PROPEXP[StormData$PROPDMGEXP == "M"] <- 1e+06
StormData$PROPEXP[StormData$PROPDMGEXP == ""] <- 1
StormData$PROPEXP[StormData$PROPDMGEXP == "B"] <- 1e+09
StormData$PROPEXP[StormData$PROPDMGEXP == "m"] <- 1e+06
StormData$PROPEXP[StormData$PROPDMGEXP == "0"] <- 1
StormData$PROPEXP[StormData$PROPDMGEXP == "5"] <- 1e+05
StormData$PROPEXP[StormData$PROPDMGEXP == "6"] <- 1e+06
StormData$PROPEXP[StormData$PROPDMGEXP == "4"] <- 10000
StormData$PROPEXP[StormData$PROPDMGEXP == "2"] <- 100
StormData$PROPEXP[StormData$PROPDMGEXP == "3"] <- 1000
StormData$PROPEXP[StormData$PROPDMGEXP == "h"] <- 100
StormData$PROPEXP[StormData$PROPDMGEXP == "7"] <- 1e+07
StormData$PROPEXP[StormData$PROPDMGEXP == "H"] <- 100
StormData$PROPEXP[StormData$PROPDMGEXP == "1"] <- 10
StormData$PROPEXP[StormData$PROPDMGEXP == "8"] <- 1e+08
# Assigning 0 to invalid exponent data
StormData$PROPEXP[StormData$PROPDMGEXP == "+"] <- 0
StormData$PROPEXP[StormData$PROPDMGEXP == "-"] <- 0
StormData$PROPEXP[StormData$PROPDMGEXP == "?"] <- 0
# Calculating Property damage value in dollar amount
StormData$PROPDMGVAL <- StormData$PROPDMG * StormData$PROPEXP
```

###Estimating Crop damage by dollar amount

Estimating Crop damage by dollar amount. To prepare the data, we need to assign a dollar amount value according with the magnitude of the damage.

```{r}
#exploring Crop damage data
unique(StormData$CROPDMGEXP)
# Sorting the property exponent data
StormData$CROPEXP[StormData$CROPDMGEXP == "M"] <- 1e+06
StormData$CROPEXP[StormData$CROPDMGEXP == "K"] <- 1000
StormData$CROPEXP[StormData$CROPDMGEXP == "m"] <- 1e+06
StormData$CROPEXP[StormData$CROPDMGEXP == "B"] <- 1e+09
StormData$CROPEXP[StormData$CROPDMGEXP == "0"] <- 1
StormData$CROPEXP[StormData$CROPDMGEXP == "k"] <- 1000
StormData$CROPEXP[StormData$CROPDMGEXP == "2"] <- 100
StormData$CROPEXP[StormData$CROPDMGEXP == ""] <- 1
# give 0 to invalid exponent data, so they not count in
StormData$CROPEXP[StormData$CROPDMGEXP == "?"] <- 0
# compute the crop damage value
StormData$CROPDMGVAL <- StormData$CROPDMG * StormData$CROPEXP
```

###Grouping the data by event type

There is 985 types of events.

```{r}
# aggregate the data by event
#unique(StormData$EVTYPE)
grpfatal <- aggregate(FATALITIES ~ EVTYPE, data = StormData, FUN = sum)
grpinjury <- aggregate(INJURIES ~ EVTYPE, data = StormData, FUN = sum)
grppropdmg <- aggregate(PROPDMGVAL ~ EVTYPE, data = StormData, FUN = sum)
grpcropdmg <- aggregate(CROPDMGVAL ~ EVTYPE, data = StormData, FUN = sum)

```

##Results

###1. Across the United States, which types of events are most harmful with respect to population health?

Displaying only the first 10 event types most harmful events from the 985 event types in total.

```{r}
# get top10 event with highest fatalities
grpfatal10 <- grpfatal[order(-grpfatal$FATALITIES), ][1:10, ]
# get top10 event with highest injuries
grpinjury10 <- grpinjury[order(-grpinjury$INJURIES), ][1:10, ]
par(mfrow = c(1, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
par(las=3) # make label text perpendicular to axis
barplot(grpfatal10$FATALITIES, names.arg = grpfatal10$EVTYPE, 
    main = "Weather Events With the Top 10 Highest Fatalities", 
    ylab = "Number of fatalities", 
    col = c("red4","red","tomato3","tomato","orangered3","orangered2","orangered","orange","yellow","yellow3"),
    legend = grpfatal10$FATALITIES,
    args.legend = list(title = "NUMBERS", x = "topright", cex = .7))

par(las=3) # make label text perpendicular to axis
barplot(grpinjury10$INJURIES, las = 3, names.arg = grpinjury10$EVTYPE, 
    main = "Weather Events with the Top 10 Highest Injuries", 
    ylab = "Number of injuries", 
    col = c("navy","blue4","blue2","dodgerblue4","cornflowerblue","lightblue4","lightslateblue","deepskyblue","lightblue","lightblue1"),
    legend = grpinjury10$INJURIES,
    args.legend = list(title = "NUMBERS", x = "topright", cex = .7))
#text(x= bplt2, y=grpinjury10$INJURIES, labels=as.character(grpinjury10$INJURIES), xpd=TRUE, col="white")
```
    
According with the graphic above, the Tornado is the most harmful Weather event that causes the highest fatalities (5633) and injuries (91,346).

###2. Across the United States, which types of events have the greatest economic consequences?

Displaying only the first 10 event types most harmful events from the 985 event types in total.

```{r}
grppropdmg10 <- grppropdmg[order(-grppropdmg$PROPDMGVAL), ][1:10, ]
# get top 10 events with highest crop damage
grpcropdmg10 <- grpcropdmg[order(-grpcropdmg$CROPDMGVAL), ][1:10, ]
par(mfrow = c(1, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(grppropdmg10$PROPDMGVAL/(10^9), las = 3, names.arg = grppropdmg10$EVTYPE, 
    main = "Top 10 Events with the highest Property Damages", ylab = "Damage Cost (in $ billions)", 
    col = c("maroon4","mediumvioletred","purple4","purple3","orchid4","plum4","plum3","plum2","mistyrose","moccasin"),
    legend = grppropdmg10$PROPDMGVAL/(10^9),
    args.legend = list(title = "NUMBERS", x = "topright", cex = .7))
barplot(grpcropdmg10$CROPDMGVAL/(10^9), las = 3, names.arg = grpcropdmg10$EVTYPE, 
    main = "Top 10 Events With the highest Crop Damages", ylab = "Damage Cost (in $ billions)", 
    col = c("darkgreen","darkolivegreen","forestgreen","darkolivegreen4","green4","limegreen","mediumseagreen","mediumspringgreen","olivedrab3","olivedrab1"),
    legend = grpcropdmg10$CROPDMGVAL/(10^9),
    args.legend = list(title = "NUMBERS", x = "topright", cex = .7))
```

Flood has been the Weather event causing the major number of fatalities, meanwhile Drought caused the major number of injuries.

Displaying tornado numbers during the years

```{r}
#Just Tornado data
TornadoDF <- StormData[StormData$EVTYPE == "TORNADO",]
TorFatalPerYear <- aggregate(FATALITIES ~ YEAR, data = TornadoDF, FUN = sum)
grpFatalxYear5 <- TorFatalPerYear[order(-TorFatalPerYear$YEAR), ][1:5, ]
grpFatalxYear5 <- grpFatalxYear5[order(grpFatalxYear5$YEAR),]

TorInjurPerYear <- aggregate(INJURIES ~ YEAR, data = TornadoDF, FUN = sum)
grpInjurxYear5 <- TorInjurPerYear[order(-TorInjurPerYear$YEAR), ][1:5, ]
grpInjurxYear5 <- grpInjurxYear5[order(grpInjurxYear5$YEAR),]

grppropyear <- aggregate(PROPDMGVAL ~ YEAR, data = TornadoDF, FUN = sum)
grpcropyear <- aggregate(CROPDMGVAL ~ YEAR, data = TornadoDF, FUN = sum)
grppropyear5 <- grppropyear[order(-grppropyear$PROPDMGVAL), ][1:5, ]
grppropyear5 <- grppropyear5[order(grppropyear5$YEAR),]
grpcropyear5 <- grpcropyear[order(-grpcropyear$CROPDMGVAL), ][1:5, ]
grpcropyear5 <- grpcropyear5[order(grpcropyear5$YEAR),]

par(mfrow = c(2, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
#Tornado plots
barplot(grpFatalxYear5$FATALITIES, las = 3, names.arg = grpFatalxYear5$YEAR, 
    main = "Tornado fatalities per Year", ylab = "Number of fatalities", 
    col = c("red"),
    legend = grpFatalxYear5$FATALITIES,
    args.legend = list(title = "NUMBERS", x = "topleft", cex = .7))
barplot(grpInjurxYear5$INJURIES, las = 3, names.arg = grpInjurxYear5$YEAR, 
    main = "Tornado injuries per Year", ylab = "Number of injuries", 
    col = c("blue"),
    legend = grpInjurxYear5$INJURIES,
    args.legend = list(title = "NUMBERS", x = "topleft", cex = .7))
#dollar
barplot(grppropyear5$PROPDMGVAL/(10^9), las = 3, names.arg = grppropyear5$YEAR, 
    main = "Tornado Property damages", ylab = "Damage Cost (in $ billions)", 
    col = c("maroon4","mediumvioletred","purple4","purple3","orchid4"))
barplot(grpcropyear5$CROPDMGVAL/(10^9), las = 3, names.arg = grpcropyear5$YEAR, 
    main = "Tornado Crop Damages", ylab = "Damage Cost (in $ billions)", 
    col = c("darkgreen","darkolivegreen","forestgreen","darkolivegreen4","green4"))

```